---
title: "Marathon Training Plan Data Analysis"
#author: "Steve Newns"
#date: "February 24, 2017"
output:
  github_document: default
  html_document: default
  word_document: default
---

# *Initial Data Inspection*

First, I load in the CSV file created in the **R_Cleaning_Garmin_Strava** script file, which contains the combined dataset of runs from my Philadelphia Marathon training plan from both Strava and Garmin Connect. I then inspect the data to make sure I have all the correct data in the correct data types.

```{r load dataset, echo = FALSE}
runs <- read.csv("cleanedMarathonTrainingData.csv")
str(runs)
```

Well, to start off, I have a column which seems to contain the row numbers from the CSv, so let's remove that before going back to the inspection.

```{r remove 1st col }
runs$X <- NULL
head(runs,2)
```

Alright, cool. But it looks like my Date field, **Date**, total run time field, **Time**, and average pace (min/mile) field, **Avg.Pace**, are not in the POSIXct format needed. So let's put them back into the correct format.

```{r fix POSIXct}
library(lubridate)

runs$Time <- as.POSIXct(runs$Time)#, format = '%H:%M:%S')
runs$Avg.Pace <- as.POSIXct(runs$Avg.Pace)#, format = '%H:%M:%S')
runs$Date <- as.POSIXct(runs$Date)
runs$StartTime <- hour(strptime(runs$StartTime, format = '%H:%M'))

str(runs)
```

Looks like **Month** was also not loaded as on *ordered* factor like I need it. As it is now, if I used it in my graphs, it would be ordereded alphabetically, not chronologically like I need it. So let's convert it back to an ordered factor with the correctly-specified ordering via the **levels** argument.

```{r fix month}
runs$Month <- factor(runs$Month, ordered = TRUE, levels = c("Jul","Aug","Sep","Oct","Nov"))
```

Now its time to acually make some plots. I'll start off with a simple histogram of how many miles I ran in each of my runs.

```{r plotting, echo = FALSE}
library(ggplot2)
library(RColorBrewer)

#distance histogram
ggplot(data = runs, aes(x = Distance)) + 
  geom_histogram(binwidth = 2, aes(fill = ..count..), colour = "black", boundary = 2) + 
  xlab("Distance (mi)") + 
  ylab("Frequency") + 
  ggtitle("Distribution of Miles Ran in All Runs") + 
  guides(fill=FALSE)
  
```

It looks like the majority of my runs (20 of 'em) were between 10 and 12 miles, with the next highest frequency being from 4 to 6 miles, which were mostly recovery runs, showing that there was indeed a lot decent amount of time for recovery in this plan. We can also see my one 24 mile run all the way to the right, my longest Long Run of the plan.

Now let's look at the total number of miles ran each month, with a pretty obvious expectation to see the total rise month-by-month.

```{r miles by month, echo = FALSE} 
#miles by month bars
ggplot(data = runs, aes(x = Month, y = Distance, fill = Month)) + 
  geom_bar(stat="identity") +
  xlab("Month") + 
  ylab("Total Miles") + 
  guides(fill=FALSE) +
  ggtitle("Total Miles by Month")

tapply(runs$Distance,runs$Month,sum)
```

The largest number of miles ran, 308 miles, was in September, and then we see a *ever-so-slightly* drop to 304 miles in October. This could be possibly be so if my peak mileage week was in September. We then see a large drop from October to November, which was expected due tapering up the race and the fact that the race was on November 20th, so there's no more run data after that date.

```{r check max run} 
runs[which.max(runs$Distance[runs$Date < "2016-11-20"]),]
```

Welp, I was half wrong about that I guess. It seems peak week was actually both in September *and* October, with the final run of the week coinciding with the start of October, so that was where my total mileage started to decreased *just* a bit.

Another important metric for runners is their **cadence**, which is the number of steps taken per minute. The general consensus for a good cadence to run efficiently seems to be anywhere from 160 to 180 steps per minute, with more advanced and elite runners leaning towards the higher end.


```{r cadence by month, echo = FALSE} 
#avg. cadence grouped by month
ggplot(data = runs, aes(Month, Cad)) + 
  geom_jitter(aes(colour = Month)) +
  geom_boxplot(aes(fill = Month), 
               outlier.colour = "black", alpha = 0.5) + #geom_jitter()
  xlab("Month") + 
  ylab("Cadence") + 
  guides(fill=FALSE) +
  ggtitle("Cadence per Month") 
```

So, barring some outliers, I can see that my cadence tended to generally increases month over month, and ended up averagin between 170 and 180, barring those outliers. I would assume they were mainly recovery runs, as I was mainly concerned with keeping my heart rate down so I could actually recover on them than I was with making sure my cadence was optimal. I could investigate to see if this was due to having more speed workouts as the plan progressed, in which I would assume my cadence was typically higher), or could it have been due to improved form from an ever-increasing volume????

Another metric would be my average heart rate during runs over the course of the training. Compared to the rise of my cadence, we would expect that my average heart rate would have decreased over time, as my fitness would have increased due to higher mileage and an increasing number of workouts.

```{r HR by month boxplot, echo = FALSE}
#avg. HR grouped by month

ggplot(data = runs, aes(Month, Avg.HR)) + 
  geom_jitter(aes(colour = Month)) +
  geom_boxplot(aes(fill = Month), 
               outlier.colour = "black", alpha = 0.5) + #geom_jitter()
  xlab("Month") + 
  ylab("Avg. Heart Rate") + 
  ggtitle("Average Heart Rate per Month") 
```

So we can see that my hear rate had generally decreased month over month. Now while I would hope it was due to increased fitness (and I still believe that to be the case, just by how I felt during my runs), the drops in temperature as we moved from summer to fall also may have played a part. Since it would have been cooler out, the body would not have been working as hard to cool itself down during my runs, so my heart rate would not have been as high. Maybe finding the historical temperature data and bringing it in could bring some insights into this question.

Now let's look at cadence and heart rate by Run Type as well for comparisons.

```{r cadence by run type, echo = FALSE} 
#avg. cadence grouped by month
ggplot(data = runs, aes(RunType, Cad)) + 
  geom_jitter(aes(colour = RunType)) +
  geom_boxplot(aes(fill = RunType), 
               outlier.colour = "black", alpha = 0.5) + #geom_jitter()
  xlab("Run Type") + 
  ylab("Cadence") + 
  guides(fill=FALSE) +
  coord_flip() + 
  ggtitle("Cadence per Run Type") 
```

We can ignore the variability in the **Race** run type, since it's only on race. But we can still see that it was either my maximum cadence, or the second or third highest after those 2 quickest workout cadences. We can also see that my recovery runs had the least variability, and recovery runs as well as normal runs. I also expected my workouts to have the highest cadence, but it has the lowest median. I can only assume that this was due to recovery sections of interval runs, or warm-ups and cooldowns during tempo runs which were *much* slower and relaxed than even recovery runs. Now let's look at heart rate.

```{r heart rate by run type, echo = FALSE}
#avg. cadence grouped by month
ggplot(data = runs, aes(RunType, Avg.HR)) + 
  geom_jitter(aes(colour = RunType)) +
  geom_boxplot(aes(fill = RunType), 
               outlier.colour = "black", alpha = 0.5) +
  xlab("Run Type") + 
  ylab("Avg Heart Rate") + 
  guides(fill=FALSE) +
  ggtitle("Average Heart Rate per Run Type") 
```

As expected, my heart rate was lowest during recovery runs. We can also see that there were 2 runs were my heart rate monitor was off, since my average heart rate was around 70. Also, my average heart rate was not very variable for the other 3 types of runs, excluding the race. My Long runs also tended to have a lower average heart rate than my "normal" runs, which I personally did not expect. But this does a make a little sense, since I was supposed to run at a bit of a slower pace so that I could run longer. Also, I did expect my heart rate to be higher during workouts, since they were designed to to push my heart rate to the max during **V02 intervals**, and was supposed to be elevated during those "comfortably hard" **tempo runs**. I can also see that there was one single workout wherein it was higher than my average heart rate during the marathon.

```{r max heart rate}
runs[which.max(runs$Avg.HR),]
```

So we can see that my max heart rate was during my 10k Tune-up Race, which does not surprise me since it SUCKED.

```{ r heart rate over time} 
ggplot(data = runs, aes(x = Date, y = Avg.HR)) + 
  geom_line(aes(group=1)) +
  theme(axis.text.x = element_blank()) +
  xlab("Time") + 
  ylab("Average Heart Rate") + 
  ggtitle("Average Heart Rate Over Time") + 
  labs(caption="*Very slight general decrease, with outliers of about 100 in the 1st third of the plan and 
70 in the last 3rd of the plan")
```

#average pace over plan line
```{r avg pace}
ggplot(data = runs, aes(x = Date, y = Avg.Pace)) + 
  geom_point() + # by color? 
  #?stat_smooth() + 
  theme(axis.text.x = element_blank()) +
  xlab("Time") + 
  ylab("Average Heart Rate") + 
  ggtitle("Average Heart Rate Over Time") + 
  labs(caption="*Very slight general decrease, but not linear")
```
